INGRID: Creating Languages in MPS from ANTLR Grammars

### 2016-11-05 chapters/evaluation.tex
evaluation - done

% TODO continue from here

\paragraph{C\#}.
Pred generovanim MPS jazyka pro C\# bylo nutne upravit gramatiku (coz zabralo asi hodinu prace) aby INGRID vyprodukoval pomerne rozumny vysledek.
The definition of C\# in MPS is quite large and complex.
Obsahuje asi vic nez 800 konceptu (zkontrolovat).
Import of the adjusted C\# language funguje docela hezky (ale je hodne velky a dlouho se kompiluje).
Ale na skutecne dobre pouzitelny jazyk C\# uvnitr MPS by to chtelo jeste vic manualnich uprav gramatiky a take urcitych aspektu vysledneho MPS jazyka (editor, textgen, apod).

\paragraph{Other languages.}
We also experimented with import of other languages, such as Python and Ruby.
Import jazyku Python a Ruby moc nefunguje protoze gramatika je napsana stylem ktery neni vhodny pro INGRID.
INGRID cannot handle cyclic rules (which could be found in Python grammar, for example).
Cyklicka pravidla to tedy neumi, jen na ne upozorni (a rekne, ktere pravidlo je obsahuje).
Museli bychom gramatiku rucne upravit (bude to ale nase future work).

\paragraph{General observations.}
Most of the languages (that we tried to define in MPS, use/import into MPS) are quite complex when it comes to their structure and the overall syntax variety, so defining them manually would be a very tedious, error-prone and time-consuming process.
\todo{Discuss where INGRID helps the most based on the experience.}
(Our experience/experiments show that) Automated generation of the Structure aspect is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
On the other hand, we found out that manual tweaking/adjusting of the Editor and Textgen aspects (adding line breaks and indentation) is very effective, and takes only a relatively little/short time (much less than we expected).
The imported language can be, in our opinion, adjusted very fast into a more readable form.
The generated MPS language can be customized very fast into a very usable form.
In fact/total, the adjusted version of each language was created from the original import by the first author, who spent no more than between 20 to 60 minutes of working on each of them.
The first author spent only a couple of minutes of (manually) adjusting the languages after the automated import.
Our experience with all the languages confirms our hypothesis (that code layout in the Editor aspect can be adjusted very fast, from the basic default single-row layout), and justifies some of the decisions we have described in Section X (editor aspect) and Section Y (textgen), proving that our approach is quite practical.
It takes a very short time to reach optimal results (because MPS allows doing this very efficiently), as we show on the example of importing the JavaScript language and manually adjusting all editors that needed it, all that in a less than hour time.
\todo{Mozna rict:} However, during our work on this project, we also tried to design some/various automated heuristics, but all that we came up with so far give rather suboptimal/bad results especially when compared to results produced by a human user.

>>>>

\paragraph{C\#.}
Another very complex programming language that we used for experiments is C\#.
Before we could run INGRID, we had to manually adjust the ANTLR grammar of C\# in order to ensure that INGRID produces a reasonably useful MPS language.
Roughly one hour of manual effort was needed for this step.
Nevertheless, additional modifications of the grammar and selected aspects of the MPS language (Editor, TextGen) are still needed to get an optimal result.

The generated definition of C\$ in MPS is quite large, involving more than 800 concepts, and therefore building of the language takes a rather long time.

\paragraph{Other languages.}
We also tried to apply INGRID on few other languages, such as Python and Ruby.
Results are mixed because ANTLR grammars of these languages are written in a style that is not fully compatible with INGRID.
For example, INGRID cannot handle cyclic rules (which appear in the Python grammar) and just prints a warning on the screen.
Significant modifications of the grammar would be necessary.
We plan to address this limitation in the future.

\paragraph{General observations.}
The main overall benefit of the INGRID method is partial automation.
Most of the languages discussed above, which we tried to create in MPS by using INGRID, are quite complex regarding their structure and syntax variety.
Therefore, completely manual definition would be a very time-consuming and error-prone process.
Fully automated generation of the Structure aspect is where INGRID spares the user from many hours of tedious and sometimes quite challenging work.

On the other hand, our experiments with complex languages show that, in the case of the Editor and Textgen aspects, manual adjustment (e.g., adding line breaks and indentation) is a very effective approach that takes only a short time --- in particular, much less time than we expected.
The first author spent between 20 and 60 minutes of work on each language, when adjusting the result of automated generation into a more useful and readable form.
MPS IDE provides good support for efficient tweaking of the code layout in Editor and TextGen, allowing users to reach optimal results very fast.
All of this justifies some of our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
Despite that, we also tried to design some automated heuristics during our work on the INGRID project, but so far all yield rather average results especially when compared to what human users can achieve efficiently.
